---
title: "Data Summary in R"
author: "Sheherazade (R ladies Bogor)"
date: "02/07/2019"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before!

+ Always check which project and working directory you are working now
```{r}
getwd()
```

+ All the dataset is in the folder used as a working directory

+ All the materials are compiled from different sources:

  - McGarigal, K. Course Analysis of Environmental Data: Introduction to R. University of Massachusetts Amherst. www.umass.edu/landeco/teaching/ecodata/syllabus.ecodata.pdf

  - www.statisticssolutions.com/data-levels-of-measurement/

  - www.statmethods.net/stats/descriptives.html

  - www.r-bloggers.com/r-tutorial-series-summary-and-descriptive-statistics/

  - www.sthda.com/english/wiki/descriptive-statistics-and-graphics


# Data Summary in R

Variables -> named computer memory. For example, a list of species recorded stored in variable named 'species_name'.

```{r}
pi <- 3.14 #real
number_species <- 327 #integer
species_name <- 'Panthera tigris' #character/string, should be entered in quotes
rainforest <- TRUE #logical (only TRUE or FALSE)
```

## Data Structures

1. **vectors** : one-dimensional, a single data type (*e.g.* integer, real, and character/string).

```{r}
example_vector<-c(2,4,6,8,10)
example_vector
(example_vector<-c(2,4,6,8,10))
example_vector[2]     #individual item is identified by the sequence (1-n)
```

2. **matrices** : two dimensional, created by binding two or more vectors of the same type and length.

```{r}
example_vector_2<-c(1,3,5,7,9)    
example_matrix<-cbind(example_vector,example_vector_2)
example_matrix

example_matrix[4,2]   #row 4, column 2
example_matrix[5,]    #row 5
example_matrix[,1]    #column 1
example_matrix[]      #all
example_matrix[,1:2]   #column 1 to 2
example_matrix[,-1]   #all except column 1. minus artinya kecuali
```

3. **data frames** : one to multi-dimensional. Similar to spreadsheet. Column - field.

```{r}
data1<-read.csv("data1.csv")
head(data1)
```

4. **lists** : compound objects of associated data. It could consist of vector, matrix, and data frames.

```{r}
example_list<-list(example_vector,data1)
example_list
names(example_list)<-c('vector','data frame')
example_list
```

## Data types

Understanding classification of data is a prerequisite for data summary and visualization, including to determine the correct statistical analysis. There are four levels of data measurement:

**1. Nominal data**  
Labels for a particular category, usually referred as categorical data. For example:

+ Gender: male and female
+ Religion: moslem, christian, hindu, budha, etc.
+ Occupation: fisherman, businessman, etc.

In R, categorical data is represented as factors that contain levels.

You can check the structure of data using 'str' function. 

```{r}
data1<-read.csv("data1.csv")
names(data1)
str(data1$gender)
```

Gender is a factor, with two levels: male and female.

**2. Ordinal data**  
This is similar to nominal, but it has orders. For example:

+ Level of income: lower class, middle class, and upper class
+ Level of agreement: strongly disagree, disagree, neutral, agree, and strongly agree

In R, function `factor()` assigns order to nominal data.

```{r}
factor_income<-factor(data1$income,order=TRUE,levels=c("lower_class","middle_class","upper_class"))
str(factor_income)
```

Nominal and ordinal data are also reffered to as qualitative data.

**3. Interval data**  

+ has values of equal intervals that mean something (20-30$^\circ$C and 40-50 $^\circ$C have 10 degree differences)
+ no absolute zero

For example: Temperature (Celcius, Fahrenheit)

**4. Ratio data**

+ like interval data but has absolute zero

For example: height, weight (no 0 kg, but there is 0 $^\circ$C)

Interval and ratio data are usually called as quantitative data, which may be classified as **discrete** (integer in R) and **continuous** (real in R). 

## Why is this important?

+ t.test -> continuous data (sugar concentration in tomatoes)
+ ANOVA -> continuous data (mice weight compared among control, treatment 1, and treatment 2)
+ linear modelling -> determine distribution (binary categorical -> binomial, count discrete -> poisson)

**So, understand your data first!**

## Other basic operations in R

### Missing values

```{r}
example<-c(1,3,5,7,9,11)     
#delete number 9 or replace it with NA (if delete 9 does not work) for this set of codes
example<-c(1,3,5,7,NA,11)     
example<-as.numeric(example)
str(example)
str(example)
is.na(example)     #to identify missing values
example_2<-example[!is.na(example)]  #to use all of the vector except the missing values
(example_2<-example[!is.na(example)]) #or
example_3<-na.omit(example)
example_3
(mean<-mean(example))
(mean_1<-mean(example,na.rm=TRUE))
```


### Creating subsets

```{r, eval = FALSE}
head(data1)
data_new<-data1[,5:7]  #select all rows and columns 5 through 7
head(data_new)

```

